### 2022-10-19 draft/draft.tex
almost done

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\setlength{\parindent}{20.0pt}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}
\newcommand{\SBM}{\mathrm{SBM}}
\newcommand{\DCBM}{\mathrm{DCBM}}

@@

@@

\hypertarget{sec:grdpg}{%
\subsection{Random Dot Product Graphs and Generalized Random Dot Product Graphs}\label{sec:grdpg}}

@@

@@

@@

@@

\begin{example}
Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \RDPG(X)$. Figure~\ref{fig:two-curves} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the adjacency spectral embedding of the graph. 
Although the community structure is not obvious from the graph, the embedding shows a clear separation between the two communities.

@@

\caption{Manifold block model described in Example 1. The latent configuration is on the left, a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.}\label{fig:two-curves}
\end{figure}
\end{example}

@@

\hypertarget{sec:nonintersect}{%
\subsection{Algorithms for Nonintersecting Manifolds}\label{sec:nonintersect}}

@@

\begin{theorem}[Community detection for the GRDPG for which the communities come from nonintersecting curves]
\label{theorem:nonintersect-curves}
\end{theorem}

\hypertarget{sec:intersect}{%
\subsection{Algorithms for Intersecting Manifolds}\label{sec:intersect}}

\hypertarget{examples-and-simulations}{%
\subsection{Examples and Simulations}\label{examples-and-simulations}}

>>>>

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\setlength{\parindent}{20.0pt}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}
\newcommand{\SBM}{\mathrm{SBM}}
\newcommand{\DCBM}{\mathrm{DCBM}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\MBM}{\mathrm{MBM}}

@@

The PABM, DCBM, and SBM are nested models, that is, the SBM is a special case of the DCBM which in turn is a special case of the SBM.
In this section, these connections are made explicit, and we explore the implications for modeling.
For ease of notation, throughout this section the rows and columns of \(P\) are assumed to be organized by community in increasing order, i.e., the first \(n_1\) rows and columns of \(P\) correspond to community \(1\), the next \(n_2\) rows and columns correspond to community \(2\), etc.
We use the notation \(P^{(k \ell)} = (P^{(\ell k)})^\top\) to denote the block of \(P\) that corresponds to the edges between communities \(k\) and \(\ell\).
In section \ref{sec:pabm-grdpg}, we will deal with this notation and convention more carefully.
For additional details on the connections among the SBM, DCBM, and PABM, refer to \citet{Noroozi2022}.

To see that the SBM is a special case of the DCBM, we can simply restrict the degree parameters \(\omega_i\) to all be a constant value, e.g., \(\omega_i \equiv 1\).
But in order to better understand the relationship between these two models, we will break down the models further.

Define \(B\) as the \(K \times K\) matrix of edge probabilities between pairs of communities in the SBM.
Then \(B_{k \ell} = \theta_{k \ell}\) and \(P_{ij} = \theta_{z_i, z_j} = B_{z_i, z_j}\).
We can further show that block \(P^{(k \ell)}\) can be written as \(P^{(k \ell)} = e^{(k)} B_{k \ell} (e^{(\ell)})^\top\) where \(e^{(k)}\) is the \(1\)-vector of dimension \(k\).
Since \(B_{k \ell}\) is a scalar, \(P^{(k \ell)} = e^{(k)} B_{k \ell} (e^{(\ell)})^\top\) is a singular value decomposition of a rank 1 matrix and thus each block \(P^{(k \ell)}\) is rank 1.
Then we can see that the entire edge probability matrix can be written as
\[P = Z B Z^\top\]
and \(Z\) is an \(n \times K\) matrix in which \(Z_{ik} = I(z_i = k) = z_{ik}\). Thus, if \(B\) is full rank, i.e., \(\rank(B) = K\), then \(P\) is also rank \(K\).

To extend the SBM to the DCBM, we replace \(e^{(k)}\) with the vector \(\omega^{(k)}\) which contains the degree parameters of vertices in community \(k\).
Then for the DCBM, \(P^{(k \ell)} = \omega^{(k)} B_{k \ell} (\omega^{(\ell)})^\top\).
Similar to the SBM, this is a singular value decomposition of a rank 1 matrix.
The full edge probability matrix can be written as
\[P = (\Omega Z) B (\Omega Z)^\top\]
for \(\Omega = \diag(\omega)\).
Since \(\Omega\) is diagonal, we again have that \(\rank(P) = K\) (again, assuming that \(B\) is full rank).
Thus, the dimensionality of the DCBM is the same as the SBM, and to restrict the DCBM to the PABM, let \(\Omega = I\).

To extend the DCBM to the PABM, we note that the \(k \ell\)\textsuperscript{th} block of the edge probability matrix can be written as \(P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^\top\) where \(\lambda^{(k \ell)} \in \mathbb{R}^{n_k}\) is the vector of popularity parameters belonging to vertices in community \(k\) that point to community \(\ell\), i.e., \(\lambda^{(k \ell)}_i = \lambda_{i + \sum_{s=1}^{k-1} n_s, \ell}\).
Similar to how \(\omega^{(k)}\) is a sub-vector of \(\omega\)
To match this decomposition with the one for the DCBM, we can fix each \(B_{k \ell} \equiv 1\) and write \(P^{(k \ell)} = \lambda^{(k \ell)} B_{k \ell} (\lambda^{(\ell k)})^\top\).
Restricting each \(\lambda^{(k \ell)} = \omega^{(k)}\) reduces the PABM to the DCBM.
To write out the full edge probability matrix of the PABM, let \(\Lambda^{(k)} =\begin{bmatrix} \lambda^{(k1)} & \cdots & \lambda^{(kK)} \end{bmatrix}\) and \(X = \blockdiag(\Lambda^{(1)}, ..., \Lambda^{(K)})\).
Then for some \(K^2 \times K^2\) permutation matrix \(\Pi\),
\[P = X \Pi X^\top.\]
From this decomposition, we can see that unlike the SBM and DCBM, \(\rank(P) = \rank(X) = K^2\).
We will explore this decomposition of the edge probability matrix of the PABM in more detail in section \ref{sec:pabm-grdpg}.

The full hierarchy of nested block models is summarized in table \ref{tab:hierarchy}.

\begin{table}

\caption{\label{tab:hierarchy}Summary of an hierarchical or nested view of block momdels.}
\centering
\begin{tabular}[t]{l|l|l|l|l}
\hline
Model & Parameters & Dimensions & Matrix Decomposition & Reduction to simpler model\\
\hline
SBM & $K (K-1) / 2$ & $K$ & $P = Z B Z^\top$ & \\
\hline
DCBM & $K (K-1) / 2 + n$ & $K$ & $P = (\Omega Z) B (\Omega Z)^\top$ & $\omega = e$\\
\hline
PABM & $n K$ & $K^2$ & $P = X \Pi X^\top$ & $\lambda_{ik} = \omega_i \sqrt{\theta_{z_i, k}}$\\
\hline
\end{tabular}
\end{table}

@@

\hypertarget{sec:grdpg}{%
\subsection{The Generalized Random Dot Product Graph}\label{sec:grdpg}}

@@

Now that we defined the GRDPG as a generative model for Bernoulli random graphs,
we need a method for estimation.
We can use Adjacency Spectral Embedding (ASE) \citep{doi:10.1080/01621459.2012.699795} to recover the latent vectors of a GRDPG.
This procedure consists of taking the spectral decomposition of \(A\) and keeping the \(p + q\) largest eigenvalues (in modulus) and corresponding eigenvectors of \(A\).

\begin{definition}[Adjacency spectral embedding]
\label{def:ase}
Let $A$ be an $n
\times n$ adjacency matrix and denote the
eigendecomposition of $A$ as
$$A = \sum_{i} \hat{\lambda}_i \hat{v}_i \hat{v}_i^{\top}, \quad
\text{where} \,\,
|\hat{\lambda}|_1 \geq |\hat{\lambda}_2| \geq \dots \geq
|\hat{\lambda}_n|.$$
Let $\hat{D}$ be a diagonal matrix whose diagonal entries are
the eigenvalues $\{\hat{\lambda}_i\}_{i=1}^{d}$ arranged in decreasing values (not in
decreasing modulus) and let $\hat{V}$ be the $n \times d$ matrix whose
columns are the corresponding eigenvectors $\{\hat{v}_i\}_{i=1}^{d}$
arranged in the same order as the diagonal entries of $\hat{D}$. Now
define $\hat{X} = \hat{V} |\hat{D}|^{1/2}$ where the $|\cdot|$
operation is applied elementwise to the entries of $\hat{D}$. Then
$\hat{X}$ serves as an estimate of $X$, up to some non-identifiable
orthogonal transformation $Q$ as described in
remark~\ref{rem:non_identifiable} and definition~\ref{def:indefinite};
see \citet{rubindelanchy2017statistical} for further details.
\end{definition}

\citet{doi:10.1080/01621459.2012.699795} and \citet{rubindelanchy2017statistical} showed that the ASE is consistent in estimating the latent vectors, up to an indefinite orthogonal transformation.
More specifically, for sparsity parameters such that \(n \rho_n = \omega (\log^{4 c} n)\) for some constant \(c > 1\) and indefinite orthogonal transformation \(Q \in \mathbb{O}(p, q)\),
\[\max_i \|Q \hat{x}_i - x_i \| = O_P \bigg( \frac{\log^c n}{n^{1/2}} \bigg)\]
where \(x^\top\) and \(\hat{x}^\top\) are the rows of \(\hat{X}\) (from the ASE) and \(X\), respectively.

\begin{example}
Suppose that latent vectors for an RDPG are drawn from the curve 
$x(t) = \begin{bmatrix} t^2 & 2 t (t-1) \end{bmatrix}^\top$, $t \in [0, 1]$. 
Let $x_i = x(t_i)$. 
Then for each $i < j$, the edges are drawn as $A_{ij} \indep \Bernoulli(x_i^\top x_j)$. 
If after observing $A$, we wish to estimate the curve, the ASE provides an estimator in which the estimated latent vectors converge to the curve, up to an orthogonal transformation (figure \ref{fig:ase-example}).

{\centering \includegraphics{draft_files/figure-latex/ase-example-1}

\caption{200 points drawn from a curve (left), and the ASE of an adjacency matrix drawn from this curve as an RDPG (right). The red line is the original curve rotated to match the ASE.}\label{fig:ase-example}
\end{figure}

\end{example}

@@

For each simulation, community labels are drawn from a multinomial
distribution, the popularity vectors \(\{\lambda^{(k \ell)}\}_K\) are drawn
from two types of joint distributions depending on whether \(k = \ell\) or \(k \not = \ell\). The edge probability matrix \(P\) is constructed using the popularity
vectors and finally the adjacency matrix \(A\)
is drawn \(A \sim \mathrm{Bernoulli}(P)\). OSC (algorithm \ref{alg:osc}) is then used for community detection, and this
method is compared against (1) SSC using the spectral embedding \(\hat{V}\)
(algorithm \ref{alg:ssc}), (2) SSC using the rows of the
observed adjacency matrix \(A\) as is done in \citet{noroozi2019estimation}
and (3) modularity maximization (MM) as is done in
\citet{307cbeb9b1be48299388437423d94bf1}. We denote the two SSC
implementations using the rows of \(A\) and using the spectral embedding
of \(A\) as SSC-A and SSC-ASE, respectively.
The parameters \(\vartheta\)
that controls the sparsity for SSC-A and SSC-ASE were chosen via a preliminary cross-validation experiment.
In practice, \(\vartheta\) that guarantee SDP (if it is possible for the particular simulated data) often result in more than \(K\) disconnected subgraphs, so a smaller \(\vartheta\) that does not guarantee SDP was chosen,
and the final clustering step of SSC-A and SSC-ASE was done
by fitting a Gaussian mixture model to the normalized Laplacian
eigenmap embeddings \citep{belkin03:_laplac} of the affinity matrix \(B\).
We also estimate the latent popularity vectors \(\{\lambda^{(k \ell)}\}\)
by assuming that the true community labels are known and then apply
algorithm \ref{alg:param_est}, and we compare this estimation method against an
MLE-based estimator as described in \citet{noroozi2019estimation} and
\citet{307cbeb9b1be48299388437423d94bf1}.

Modularity maximization is NP-hard, so
\citet{307cbeb9b1be48299388437423d94bf1} used the extreme points
(EP) algorithm (\cite{le2016}) as a greedy
relaxation of the optimization problem; the EP algorithm has a running
time of \(O(n^{K-1})\) where \(n\) is the number of vertices in the graph
and \(K\) is the number of communities.
For these simulations we instead replace the EP algorithm with the
Louvain algorithm for modularity maximization,
as the implementation of the EP algorithm in
\citet{307cbeb9b1be48299388437423d94bf1} is too computationally expensive for \(K > 2\). For \(K = 2\), it
was verified that the Louvain algorithm produces comparable results
to EP-MM.

For comparing methods, we define the community detection error as:
\[L_c(\hat{\sigma}, \sigma; V) = \min_\pi \sum_i I(\pi \circ \hat{\sigma}(v_i) = \sigma(v_i))\]
where \(\sigma(v_i)\) is the true community label of vertex \(v_i\), \(\hat{\sigma}(v_i)\) is the predicted label of \(v_i\), and \(\pi\) is a permutation operator.
This is effectively the ``misclustering count'' of clustering function \(\hat{\sigma}\).

For parameter estimation, because the popularity parameters \(\{\lambda_{ik}\}\) are unidentifiable, we instead estimate the edge probabilities \(P_{ij} = \lambda_{i z_j} \lambda_{j z_i}\) via the quantities \(\hat{P}_{ij} = \hat{\lambda}_{iz_j} \hat{\lambda}_{jz_i}\).
The parameter estimation error is then given by the normalized Frobenius norm of \(P\) divided by the number of vertices, i.e.,
\[\mathrm{RMSE}(\hat{P}, P) = \frac{1}{n} \|\hat{P} - P\|_F.\]

We also note that unlike the MLE-based method \citep{307cbeb9b1be48299388437423d94bf1}, the ASE method in algorithm \ref{alg:param_est} can be trivially modified so as to not require the community labels if we are only interested in estimating \(P\).
More specifically we first compute the ASE \(\hat{Z}\) of \(A\) (see definition \ref{def:ase}) and then compute \(\hat{P} = \hat{Z} I_{p,q} \hat{Z}^{\top}\).
The resulting estimate \(\hat{P}\) will have the same convergence rate as that given in equation (\ref{eq:conv_prob}).

\hypertarget{sec:balanced-communities}{%
\subsubsection{Balanced Communities}\label{sec:balanced-communities}}

In each simulation, community labels \(z_1, ..., z_n\) were drawn from a
multinomial distribution with mixture parameters
\(\{\alpha_1, ..., \alpha_K\}\), then \(\{\lambda^{(k \ell)}\}_K\) according
to the drawn community labels, \(P\) was constructed using the drawn
\(\{\lambda^{(k \ell)}\}_K\), and \(A\) was drawn from \(P\).

For these examples, we set the following parameters:

\begin{itemize}
\tightlist
\item
  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
\item
  Number of underlying communities \(K = 2, 3, 4\)
\item
  Mixture parameters \(\alpha_k = 1 / K\) for \(k = 1, ..., K\), (i.e.,
  each community label has an equal probability of being drawn)
\item
  Community labels
  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
\item
  Within-group popularities
  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\)
\item
  Between-group popularities
  \(\lambda^{(k \ell)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\) for
  \(k \neq \ell\)
\end{itemize}

Fifty simulations were performed for each combination of \(n\) and \(K\).
The results for community recovery and parameter estimations are presented in figures \ref{fig:clust_err_ct_sim} and \ref{fig:p_block_est}, respectively.

Figure \ref{fig:clust_err_ct_sim} shows that OSC recovers the community
perfectly as \(n\) increases, i.e., the number of mislabeled vertices
goes to \(0\).
The performance of SSC-ASE is comparable to OSC for \(K \geq 3\) but is noticeably worse when \(K = 2\).
Similarly, SSC on both the embedding and on the adjacency matrix produces similar trends for \(K > 2\).
The difference in performance between SSC-A and SSC-ASE for \(K = 2\) can be attributed to the final spectral clustering step of the affinity matrix. While the subspace detection property is guaranteed for large \(n\), in our simulations, setting the
sparsity parameter \(\vartheta\) to the required value usually resulted
in more than \(K\) disconnected subgraphs in the affinity matrix
\(\hat{B}\).
We instead chose a smaller sparsity parameter,
necessitating a final clustering step.
A GMM was fit to the normalized Laplacian eigenmap of \(\hat{B}\),
but visual inspection suggests that the communities are not
distributed as a mixture of Gaussians in the eigenmap.
A different choice of mixture distribution may result in better performance.

\begin{figure}[tp]
{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/clust_err_ct_sim-1}
}
\caption{Median and IQR of community detection error. Communities are approximately balanced. Simulations were repeated 50 times for each sample size.}
\label{fig:clust_err_ct_sim}
\end{figure}

Given ground truth community labels, figure \ref{fig:p_block_est} shows
that algorithm \ref{alg:param_est} and the MLE-based
plug-in estimators perform comparably, with root mean square
error decaying at rate approximately \(n^{-1/2}\) as \(n\) increases.

\begin{figure}[tp]
{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/lambda_est_p-1}}
\caption{Median and IQR RMSE for edge probability matrices
  reconstructed from the outputs of algorithm~\ref{alg:param_est}
  (red) compared against outputs of an MLE-based method (blue) proposed in \citet{307cbeb9b1be48299388437423d94bf1}.
  Simulations were repeated 50 times for each sample size. Communities were drawn to be approximately balanced.}
\label{fig:p_block_est}
\end{figure}

\hypertarget{sec:imbalanced-communities}{%
\subsubsection{Imbalanced Communities}\label{sec:imbalanced-communities}}

Simulations performed in this section are the same as those in the
previous section with the exception of the mixture parameters
\(\{\alpha_1, ..., \alpha_K\}\) used to draw community labels from the
multinomial distribution. For these examples, we set the following
parameters:

\begin{itemize}
\tightlist
\item
  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
\item
  Number of underlying communities \(K = 2, 3, 4\)
\item
  Mixture parameters \(\alpha_k = \frac{k^{-1}}{\sum_{\ell=1}^K \ell^{-1}}\)
  for \(k = 1, ..., K\)
\item
  Community labels
  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
\item
  Within-group popularities
  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\)
\item
  Between-group popularities
  \(\lambda^{(k \ell)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\) for
  \(k \neq \ell\)
\end{itemize}

Fifty simulations were performed for each combination of \(n\) and \(K\).
The results for community recovery and parameter estimations are presented in figures \ref{fig:clust_err_ct_sim_imba} and \ref{fig:lambda_est_p_imba}, respectively.

From figure \ref{fig:clust_err_ct_sim_imba} we once again see that the
number of mislabeled vertices trending to \(0\) for OSC.
The performance of SSC-ASE is comparable to that of OSC for \(K > 2\) but is worse when \(K = 2\).
Figure \ref{fig:lambda_est_p_imba} indicates that the parameter estimation error also decays at rate \(n^{-1/2}\) similar to that in the balanced communities setting.

\begin{figure}[H]
{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/clust_err_ct_sim_imba-1}
}
\caption{Median and IQR of community detection error. Communities are imbalanced. 
Simulations were repeated 50 times for each sample size.}\label{fig:clust_err_ct_sim_imba}
\end{figure}

\begin{figure}[H]
{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/lambda_est_p_imba-1}
}
\caption{Median and IQR RMSE of edge probabilities derived from the
  outputs of Algorithm \ref{alg:param_est} (red) compared against an
  MLE-based method (blue) described in
  \cite{307cbeb9b1be48299388437423d94bf1}.
  Simulations were repeated 50 times for each sample size. Communities were drawn to be imbalanced.}
\label{fig:lambda_est_p_imba}
\end{figure}

\hypertarget{sec:disassortative-models}{%
\subsubsection{Disassortative Models}\label{sec:disassortative-models}}

\citet{rubindelanchy2017statistical} demonstrated the power of applying GRDPG-based approaches to disassortative block models.
Likewise, demonstrate that GRDPG-based algorithms OSC and SSC-ASE perform well on disassortative PABMs.
Simulations performed in this section are the same as those in section \ref{sec:balanced-communities} with the exception of the distributions from which within and between-group popularity parameters are drawn.
Here, we draw these parameters such that the expected value is \(1/3\) for the within-group popularity parameters and \(2/3\) for the between-group popularity parameters:

\begin{itemize}
\tightlist
\item
  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
\item
  Number of underlying communities \(K = 2, 3, 4\)
\item
  Mixture parameters \(\alpha_k = \frac{k^{-1}}{\sum_{\ell=1}^K \ell^{-1}}\)
  for \(k = 1, ..., K\)
\item
  Community labels
  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
\item
  Within-group popularities
  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\)
\item
  Between-group popularities
  \(\lambda^{(k \ell)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\) for
  \(k \neq \ell\)
\end{itemize}

OSC, SSC-ASE, and SSC-A perform similarly on disassortative PABMs compared to the simulations in sections \ref{sec:balanced-communities} and \ref{sec:imbalanced-communities}
(figures \ref{fig:clust_k_disassortative} and \ref{fig:lambda_est_disassortative}).

\begin{figure}[H]
{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/clust_err_k_disassortative}
}
\caption{Median and IQR of community detection error. 
Communities are approximately balanced. 
Edge probabilities were drawn to be disassortative. 
Simulations were repeated 50 times for each sample size.}
\label{fig:clust_k_disassortative}
\end{figure}

\begin{figure}[H]
{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/lambda_est_disassortative}
}
\caption{Median and IQR RMSE of edge probabilities derived from the
  outputs of algorithm \ref{alg:param_est} (red) compared against an
  MLE-based method (blue) described in
  \cite{307cbeb9b1be48299388437423d94bf1}.
  Simulations were repeated 50 times for each sample size. 
  Communities were drawn to be approximately balanced.
  Edge probabilities were drawn to be disassortative.}
\label{fig:lambda_est_disassortative}
\end{figure}

\begin{remark}
While we call the models in this section ``disassortative'', it is our view that the assortative/disassortative distinction is not applicable to PABMs due to its flexibility. 
Unlike the SBM and DCBM, in the PABM, each vertex is free to have a higher affinity to its own community or to other communities, independent of each other. 
In other words, within the same community, $v_i$ may have a larger popularity parameter to its own community whereas $v_j$ may have a larger popularity parameter to a different community. 
Furthermore, when viewed as GRDPGs, the assortativity or disassortativity of SBMs and DCBMs affects whether $P$ is positive semidefinite, which affects ASE-based approaches to analyzing the graph \citep{rubindelanchy2017statistical}, 
but in the full rank PABM, $P$ is never positive semidefinite and will always have $K (K + 1) / 2$ positive eigenvalues and $K (K - 1) / 2$ negative eigenvalues. 
\end{remark}

@@

In the first example, we applied OSC (algorithm \ref{alg:osc}) to the Leeds Butterfly
dataset \citep{Wang_2018} consisting of visual similarity measurements
among 832 butterflies across 10 species. The graph was modified to match
the example from \citet{noroozi2019estimation}: Only the \(K=4\) most
frequent species were considered, and the similarities were discretized
to \(\{0, 1\}\) via thresholding.
Figure \ref{fig:butterfly} shows a sorted adjacency matrix sorted by the resultant clustering.

Comparing against the ground truth species labels, \citet{noroozi2019estimation}
report that SSC on the adjacency matrix achieves an adjusted Rand index of 73\% in their implementation, whereas OSC achieves 92\% and SSC on the ASE achieves 96\%.

{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/butterfly-1}

\caption{Adjacency matrix of the Leeds Butterfly dataset 
after sorting by the clustering outputted by OSC.}\label{fig:butterfly}
\end{figure}

In the second example (figure \ref{fig:mp} and table \ref{tab:unnamed-chunk-6}), we applied OSC to the British MPs Twitter network
\citep{greene2013producing}, the Political Blogs network
\citep{10.1145/1134271.1134277}, and the DBLP network
\citep{NIPS2009_3855, 10.1007/978-3-642-15880-3_42}.
For this data analysis, we subsetted the data as described in
\citet{307cbeb9b1be48299388437423d94bf1} for their analysis of the
same networks. Our methods slightly underperformed compared to modularity
maximization, although performance is comparable.
The run time of OSC is however much smaller than that of modularity maximization.

\begin{figure}[H]
{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/mp-1}
}
\caption{Adjacency matrices of (from left to right) the British MPs, Political Blogs, and DBLP networks after sorting by the clustering outputted by OSC.}\label{fig:mp}
\end{figure}

\begin{table}[H]
\centering
\begin{tabular}[t]{l|r|r|r}
\hline
Network & MM & SSC-ASE & OSC\\
\hline
British MPs & 0.003 & 0.012 & 0.006\\
\hline
Political Blogs & 0.050 & 0.187 & 0.062\\
\hline
DBLP & 0.028 & 0.072 & 0.059\\
\hline
\end{tabular}
\caption{\label{tab:unnamed-chunk-6}Community detection error rates on the British MPs Twitter, Political Blogs, and DBLP networks using modularity maximization, sparse subspace clustering, and OSC.}
\end{table}

In the third example (figure \ref{fig:households-figure} and
table \ref{tab:households-table}),
we analyzed the Karantaka villages data studied by
\citet{DVN/U3BIHX_2013}. We chose the \texttt{visitgo}
networks from villages 12, 31, and 46 at the household level.
In these networks, each node is a household and each edge is
an interaction between members of pairs of households.
The label of interest is the religious affiliation.
The networks were truncated to religions ``1'' and ``2'',
and vertices of degree 0 were removed.
The villages were chosen based on there being an adequate number of nodes
between households within each religion.

\begin{figure}[H]
{\centering \includegraphics{/home/johnkoo/dev/pabm-grdpg/summary_files/figure-latex/unnamed-chunk-7-1}
}
\caption{Adjacency matrix of the Karnataka villages data, arranged by the clustering produced by OSC (left). The villages studied here are, from left to right, 12, 31, and 46.}\label{fig:households-figure}
\end{figure}

\begin{table}[H]
\centering
\begin{tabular}[t]{l|r|r|r}
\hline
Network & MM & SSC-ASE & OSC\\
\hline
Village 12 & 0.270 & 0.291 & 0.227\\
\hline
Village 31 & 0.125 & 0.059 & 0.051\\
\hline
Village 46 & 0.052 & 0.069 & 0.056\\
\hline
\end{tabular}
\caption{\label{tab:households-table}Community detection error rates for identifying household religion.}
\end{table}

@@

\begin{example}
\label{ex:separated-curves}
Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \RDPG(X)$. Figure~\ref{fig:two-curves} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the ASE of the graph. 
Although the community structure is not obvious from the graph, the embedding shows a clear separation between the two communities.

@@

\caption{Manifold block model from two disjoint curves. The latent configuration is on the left, a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.}\label{fig:two-curves}
\end{figure}
\end{example}

@@

\hypertarget{sec:nonintersect}{%
\subsection{Algorithm for Nonintersecting Manifolds}\label{sec:nonintersect}}

@@

\begin{example}
\label{ex:separated-curves-2}
Refer back to the RDPG in example \ref{ex:separated-curves} in which the latent vectors are drawn uniformly from two nonintersecting curves. 
Figure \ref{fig:two-curves} shows that while the ASE introduces noise to the original latent vectors, there is still sufficient separation between the two communities. 
Applying single-linkage clustering on the ASE results in recovering the original community labels, up to permutation (figure \ref{fig:two-curves-2}).

{\centering \includegraphics{draft_files/figure-latex/two-curves-2-1}

\caption{Single-linkage clustering applied to the ASE of an MBM with two nonintersecting curves. The ASE is on the left and the dendrogram from single-linkage clustering, colored by the original community labels, is on the right.}\label{fig:two-curves-2}
\end{figure}
\end{example}

\hypertarget{sec:intersect}{%
\subsection{Algorithm for Intersecting Manifolds}\label{sec:intersect}}

The consistency of the ASE provides asymptotic guarantees for perfect community label recovery for the nonintersecting MBM.
This relies on the fact that single-linkage clustering and \(\eta\)-neighborhood clustering graphs result in disconnected components if clusters of points are far away from each other.
However, many graphs do not behave so nicely.
For example, the DCBM and PABM can both be viewed as MBMs in which the latent vectors come from (linear) manifolds that intersect at the origin (see section \ref{sec:sbm-dcbm-grdpg}).
Nonlinear intersecting manifolds have also been observed in ASEs of real graphs \citep{SannaPassino2022, https://doi.org/10.48550/arxiv.1709.05454}.
While the DCBM and PABM cases can be handled by algorithms specific to linear structures such as SSC and OSC, nonlinear intersecting manifolds require a more flexible approach.
For the intersecting case, we restrict the MBM to the case in which each \(F_{z_i}\) is identical and univariate on the unit interval \([0, 1]\).
The full mixture model is described as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Draw \(t_1, ..., t_n \iid F\) for probability distribution \(F\) with support \([0, 1]\).
\item
  Draw \(z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)\), the community labels.
\item
  Let each \(x_i = g_{z_i}(t_i)\) be the latent vector for vertex \(v_i\), and collect the latent vectors into matrix \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
\item
  Draw \(A \sim \GRDPG_{p,q}(X; \rho_n)\).
\end{enumerate}
